\ourlongtitle

### 2021-08-23 Bernoulli/glo-2020-prel.tex
Corrected typos

In the case of state-space models, the identity (ii) above is a well-known result typically referred to as the \emph{backward decomposition} of the joint-smoothing distribution; still, as far as known to the authors, it has never been established in the general setting considered in the present paper, and a proof of Lemma~\ref{lem:reversibility} is hence given in the supplement (Section~\ref{sec:proof:lem:reversibility}) for completeness. Importantly, as noted in \cite{cappe:2009}, the functions $(\tstat{n} h_n)_{n \in \nsetpos}$ can be expressed recursively through  
\begin{equation} \label{eq:forward:smoothing}
\tstat{n + 1} h_{n + 1}(x_{n + 1}) = \int (\tstat{n} h_n(x_n) + \addf{n}(x_n, x_{n + 1})) \, \bkw{n}(x_{n + 1}, \rmd x_n), \quad n \in \nset, 
\end{equation}
with, by convention, $\tstat{0} h_0 \equiv 0$. Here the backward kernel $\bkw{n}$ depends on the marginal $\post{n}$; thus, the recursion is driven by the marginal flow $(\post{n})_{n \in \nset}$, which may again be expressed recursively through \eqref{eq:recursion:FK:marg}. However, as these marginals are, as already mentioned, generally intractable, exact computations need typically to be replaced by approximations. The authors of \cite{delmoral:doucet:singh:2010} propose to approximate the values of each statistic $\tstat{n} h_n$ at a random, discrete support formed by particles. More precisely, assume again that the transition density $\ud{n}$ is tractable and, by induction, that we at time step $n$ have at hand a given particle sample $(\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N$ and a set of statistics $(\tstat[i]{n})_{i = 1}^\N$ such that $\tstat[i]{n}$ is an approximation of $\tstat{n} h_n(\epart{n}{i})$. Then, in order to propagate the statistics $(\tstat[i]{n})_{i = 1}^\N$ forward, one updates, in a first substep, the particle sample $(\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N$ recursively by forward sampling (Algorithm~\ref{alg:ideal:SMC}). After forward sampling, one replaces, in the definition \eqref{eq:def:backward:kernel} of $\bkw{n}$, $\post{n}$ by the corresponding particle approximation, yielding the updates 
\begin{equation} \label{eq:tstat:FFBSm:update}
\tstat[i]{n + 1} = \sum_{j = 1}^\N \trm{n}(i, j) (\tstat[j]{n} + \addf{n}(\epart{n}{j}, \epart{n + 1}{i})), \quad i \in \intvect{1}{\N}, 
\end{equation} 
where we have defined the transition kernel  
\begin{equation} \label{eq:def:trm}
\trm{n}(i, j) \eqdef \frac{\ewght{n}{j} \ud{n}(\epart{n}{j}, \epart{n + 1}{i})}{\sum_{j' = 1}^\N \ewght{n}{j'} \ud{n}(\epart{n}{j'}, \epart{n + 1}{i})}
\end{equation}
on $\intvect{1}{\N} \times \powerset{\intvect{1}{\N}}$. Since computing each $\tstat[i]{n}$ according to \eqref{eq:tstat:FFBSm:update} has a linear computational complexity in the number $\N$ of particles, the overall complexity this approach is \emph{quadratic} in $\N$. In order to deal with this significant computational burden, the authors of \cite{olsson:westerborn:2014b} suggest replacing summation by additional Monte Carlo simulation. More precisely, by sampling, for each $i$, $\K \in \nsetpos$ independent indices $(J^{(i, j)})_{j = 1}^\M$ from $\trm{n}(i, \cdot)$ and replacing \eqref{eq:tstat:FFBSm:update} by  
\begin{equation} \label{eq:PaRIS:update}
\tstat[i]{n + 1} = \frac{1}{\K} \sum_{j = 1}^{\K} \left( \tstat[\bi{n + 1}{i}{j}]{n} + \addf{n}(\epart{n}{\bi{n + 1}{i}{j}}, \epart{n + 1}{i}) \right), \quad i \in \intvect{1}{\N}, 
\end{equation}
the computational complexity can, as we shall soon see, be reduced significantly. At each iteration, the self-normalised estimator $\sumwght{n}^{-1} \sum_{i = 1}^\N \ewght{n}{i} \tstat[i]{n}$ serves as an estimator of the quantity $\post{n} \tstat{n} h_n = \post{0:n} h_n$ of interest. This second operation, which we will refer to as \emph{backward sampling}, 
$$
    (\tstat[i]{n + 1})_{i = 1}^\N \sim \mathsf{BS}((\epart{n}{i}, \tstat[i]{n}, \ewght{n}{i})_{i = 1}^\N, (\epart{n + 1}{i})_{i = 1}^\N),  
$$
is summarised in Algorithm~\ref{alg:ideal:BS}.

@@

Let us examine closer the sampling step on Line~3 in Algorithm~\ref{alg:ideal:BS}. 
In order to keep the algorithmic complexity at a reasonable level, the computation of the normalising constant of $\trm{n}(i, \cdot)$, which consists of $\N$ terms, should be avoided; otherwise, the overall complexity remains quadratic in $\N$. This is possible using, \eg,   
\begin{itemize}
\item[--] \emph{rejection sampling}. This approach relies on the mild assumption that there exists some measurable function $c$ on $\Xset_{n + 1}$ such that $\ud{n}(x_n, x_{n + 1}) \leq c(x_{n + 1})$ for all $x_{n: n + 1} \in \Xset_n \times \Xset_{n + 1}$. Then, following \cite{douc:garivier:moulines:olsson:2010}, $\trm{n}(i, \cdot)$ can be sampled from by generating a candidate $J^\ast$ from $\cat(\{ \ewght{n}{\ell}\}_{\ell = 1}^\N)$ and accepting the same with probability 
\begin{equation} \label{eq:std:acc:prob:backward:sampling}
\accprob \eqdef \frac{\ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i})}{c(\epart{n + 1}{i})}. 
\end{equation}
The procedure is repeated until acceptance, conditionally to which $J^\ast$ is distributed according to $\trm{n}(i, \cdot)$. Since the $\cat(\{ \ewght{n}{\ell}\}_{\ell = 1}^\N)$ distribution is independent of $i$, this circumvents the need to compute a normalising sum for every $i$. The approach may reduce significantly the computational complexity; indeed, as shown in \cite[Proposition~2]{douc:garivier:moulines:olsson:2010}, the expected overall complexity of the algorithm is \emph{linear} in $\N$ under certain assumptions.   
\item[--] \emph{MCMC methods}. Another possibility is to generate the variables $(\bi{n + 1}{i}{j})_{j = 1}^{\K}$ using the Metropolis-Hastings algorithm. For this purpose, let $\rho$ be some proposal transition density on $\intvect{1}{\N}^2$. Then proceeding recursively, given $\bi{n + 1}{i}{j} = J$, a candidate $J^\ast$ for $\bi{n + 1}{i}{j + 1}$ is sampled from the density $\rho(J, \cdot)$ and accepted with probability 
$$
\accprob[MH] \eqdef 1 \wedge\frac{\ewght{n}{J^\ast} \ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i}) \rho(J^\ast, J)}{\ewght{n}{J} \ud{n}(\epart{n}{J}, \epart{n + 1}{i}) \rho(J, J^\ast)}. 
$$ 
If rejection, then $\bi{n + 1}{i}{j + 1} = J$. It is close at hand is to let $\rho$ take the form of an independent proposal given by the $\cat(\{ \ewght{n}{i} \}_{i = 1}^\N)$ distribution; in that case $\accprob[MH]$ simplifies to  
\begin{equation} \label{eq:std:MH:prob:backward:sampling}
\accprob[MH] = 1 \wedge \frac{\ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i})}{\ud{n}(\epart{n}{J}, \epart{n + 1}{i})}. 
\end{equation}
With this approach, the variables $(\bi{n + 1}{i}{j})_{j = 1}^{\K}$ are conditionally dependent; this can be counteracted by including only an $m$-skeleton of this sequence in the update \eqref{eq:PaRIS:update}. An important advantage of this approach over rejection sampling is that it does not require $\ud{n}$ to be dominated.  
\end{itemize}

@@

Finally, we remind the reader that we have here considered the idealised situation where the unnormalised transition densities $(\ud{n})_{n \in \nset}$ can be evaluated pointwise, which will generally not be the case for the applications we will consider. Thus, in the next section we will approach the more general case where these transition densities are intractable but may be estimated, and we will show how consistent, asymptotically normal, and stochastically stable estimators can be produced also in such a scenario by pseudo-marginalising separately the forward and backward sampling operations.

>>>>

In the case of state-space models, the identity (ii) above is a well-known result typically referred to as the \emph{backward decomposition} of the joint-smoothing distribution; still, as far as known to the authors, it has never been established in the general setting considered in the present paper, and a proof of Lemma~\ref{lem:reversibility} is hence given in the supplement (Section~\ref{sec:proof:lem:reversibility}) for completeness. Importantly, as noted in \cite{cappe:2009}, the functions $(\tstat{n} h_n)_{n \in \nsetpos}$ can be expressed recursively through  
\begin{equation} \label{eq:forward:smoothing}
\tstat{n + 1} h_{n + 1}(x_{n + 1}) = \int (\tstat{n} h_n(x_n) + \addf{n}(x_n, x_{n + 1})) \, \bkw{n}(x_{n + 1}, \rmd x_n), \quad n \in \nset, 
\end{equation}
with, by convention, $\tstat{0} h_0 \equiv 0$. Here the backward kernel $\bkw{n}$ depends on the marginal $\post{n}$; thus, the recursion is driven by the marginal flow $(\post{n})_{n \in \nset}$, which may again be expressed recursively through \eqref{eq:recursion:FK:marg}. However, as these marginals are, as already mentioned, generally intractable, exact computations need typically to be replaced by approximations. The authors of \cite{delmoral:doucet:singh:2010} propose to approximate the values of each statistic $\tstat{n} h_n$ at a random, discrete support formed by particles. More precisely, assume again that the transition density $\ud{n}$ is tractable and, by induction, that at time step $n$ we have at hand a given particle sample $(\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N$ and a set of statistics $(\tstat[i]{n})_{i = 1}^\N$ such that $\tstat[i]{n}$ is an approximation of $\tstat{n} h_n(\epart{n}{i})$. Then, in order to propagate the statistics $(\tstat[i]{n})_{i = 1}^\N$ forward, one updates, in a first substep, the particle sample $(\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N$ recursively by forward sampling (Algorithm~\ref{alg:ideal:SMC}). After forward sampling, one replaces, in the definition \eqref{eq:def:backward:kernel} of $\bkw{n}$, $\post{n}$ by the corresponding particle approximation, yielding the updates 
\begin{equation} \label{eq:tstat:FFBSm:update}
\tstat[i]{n + 1} = \sum_{j = 1}^\N \trm{n}(i, j) (\tstat[j]{n} + \addf{n}(\epart{n}{j}, \epart{n + 1}{i})), \quad i \in \intvect{1}{\N}, 
\end{equation} 
where we have defined the transition kernel  
\begin{equation} \label{eq:def:trm}
\trm{n}(i, j) \eqdef \frac{\ewght{n}{j} \ud{n}(\epart{n}{j}, \epart{n + 1}{i})}{\sum_{j' = 1}^\N \ewght{n}{j'} \ud{n}(\epart{n}{j'}, \epart{n + 1}{i})}
\end{equation}
on $\intvect{1}{\N} \times \powerset{\intvect{1}{\N}}$. Since computing each $\tstat[i]{n}$ according to \eqref{eq:tstat:FFBSm:update} has a linear computational complexity in the number $\N$ of particles, the overall complexity this approach is \emph{quadratic} in $\N$. In order to deal with this significant computational burden, the authors of \cite{olsson:westerborn:2014b} suggest replacing summation by additional Monte Carlo simulation. More precisely, by sampling, for each $i$, $\K \in \nsetpos$ independent indices $(J^{(i, j)})_{j = 1}^\M$ from $\trm{n}(i, \cdot)$ and replacing \eqref{eq:tstat:FFBSm:update} by  
\begin{equation} \label{eq:PaRIS:update}
\tstat[i]{n + 1} = \frac{1}{\K} \sum_{j = 1}^{\K} \left( \tstat[\bi{n + 1}{i}{j}]{n} + \addf{n}(\epart{n}{\bi{n + 1}{i}{j}}, \epart{n + 1}{i}) \right), \quad i \in \intvect{1}{\N}, 
\end{equation}
the computational complexity can, as we shall soon see, be reduced significantly. At each iteration, the self-normalised estimator $\sumwght{n}^{-1} \sum_{i = 1}^\N \ewght{n}{i} \tstat[i]{n}$ serves as an estimator of the quantity $\post{n} \tstat{n} h_n = \post{0:n} h_n$ of interest. This second operation, which we will refer to as \emph{backward sampling}, 
$$
    (\tstat[i]{n + 1})_{i = 1}^\N \sim \mathsf{BS}((\epart{n}{i}, \tstat[i]{n}, \ewght{n}{i})_{i = 1}^\N, (\epart{n + 1}{i})_{i = 1}^\N),  
$$
is summarised in Algorithm~\ref{alg:ideal:BS}.

@@

Let us examine the sampling step on Line~3 in Algorithm~\ref{alg:ideal:BS} more closely. 
In order to keep the algorithmic complexity at a reasonable level, the computation of the normalising constant of $\trm{n}(i, \cdot)$, which consists of $\N$ terms, should be avoided; otherwise, the overall complexity remains quadratic in $\N$. This is possible using, \eg,   
\begin{itemize}
\item[--] \emph{rejection sampling}. This approach relies on the mild assumption that there exists some measurable function $c$ on $\Xset_{n + 1}$ such that $\ud{n}(x_n, x_{n + 1}) \leq c(x_{n + 1})$ for all $x_{n: n + 1} \in \Xset_n \times \Xset_{n + 1}$. Then, following \cite{douc:garivier:moulines:olsson:2010}, $\trm{n}(i, \cdot)$ can be sampled from by generating a candidate $J^\ast$ from $\cat(\{ \ewght{n}{\ell}\}_{\ell = 1}^\N)$ and accepting the same with probability 
\begin{equation} \label{eq:std:acc:prob:backward:sampling}
\accprob \eqdef \frac{\ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i})}{c(\epart{n + 1}{i})}. 
\end{equation}
The procedure is repeated until acceptance, conditionally on which $J^\ast$ is distributed according to $\trm{n}(i, \cdot)$. Since the $\cat(\{ \ewght{n}{\ell}\}_{\ell = 1}^\N)$ distribution is independent of $i$, this circumvents the need to compute a normalising sum for every $i$. The approach may significantly reduce the computational complexity; indeed, as shown in \cite[Proposition~2]{douc:garivier:moulines:olsson:2010}, the expected overall complexity of the algorithm is \emph{linear} in $\N$ under certain assumptions.   
\item[--] \emph{MCMC methods}. Another possibility is to generate the variables $(\bi{n + 1}{i}{j})_{j = 1}^{\K}$ using the Metropolis-Hastings algorithm. For this purpose, let $\rho$ be some proposal transition density on $\intvect{1}{\N}^2$. Then proceeding recursively, given $\bi{n + 1}{i}{j} = J$, a candidate $J^\ast$ for $\bi{n + 1}{i}{j + 1}$ is sampled from the density $\rho(J, \cdot)$ and accepted with probability 
$$
\accprob[MH] \eqdef 1 \wedge\frac{\ewght{n}{J^\ast} \ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i}) \rho(J^\ast, J)}{\ewght{n}{J} \ud{n}(\epart{n}{J}, \epart{n + 1}{i}) \rho(J, J^\ast)}. 
$$ 
If rejection, then we set $\bi{n + 1}{i}{j + 1} = J$. An appealing solution is to let $\rho$ take the form of an independent proposal given by the $\cat(\{ \ewght{n}{i} \}_{i = 1}^\N)$ distribution; in that case $\accprob[MH]$ simplifies to  
\begin{equation} \label{eq:std:MH:prob:backward:sampling}
\accprob[MH] = 1 \wedge \frac{\ud{n}(\epart{n}{J^\ast}, \epart{n + 1}{i})}{\ud{n}(\epart{n}{J}, \epart{n + 1}{i})}. 
\end{equation}
With this approach, the variables $(\bi{n + 1}{i}{j})_{j = 1}^{\K}$ are conditionally dependent; this can be counteracted by including only an $m$-skeleton of this sequence in the update \eqref{eq:PaRIS:update}. An important advantage of this approach over rejection sampling is that it does not require $\ud{n}$ to be dominated.  
\end{itemize}

@@

Finally, we remind the reader that we have here considered the idealised situation where the unnormalised transition densities $(\ud{n})_{n \in \nset}$ can be evaluated pointwise, which will generally not be the case for the applications we will consider. Thus, in the next section we will approach the more general case where these transition densities are intractable but may be estimated, and we will show how consistent, asymptotically normal, and stochastically stable estimators can be produced also in such a scenario by pseudo-marginalising the forward and backward sampling operations separately.
